---
title: "Covid-19 trends in England"
description: |
  Using ASMODEE to monitor NHS pathways in England
author:
  - name: Thibaut Jombart, Tim Taylor, Dirk Schumacher
date: "`r Sys.Date() - 1`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = "hide", fig.height = 5, fig.width = 8, dpi = 80, layout="l-body-outset")
```

## Methodology

Using the new algorithm ASMODEE (**A**utomatic **S**election of **M**odels and
**O**utlier **D**etection for **E**pidemics) we monitor changes in potential
COVID-19 cases reported through the [NHS
pathways](https://digital.nhs.uk/dashboards/nhs-pathways), including **only data
reported through 111-online**. Since the online system dispatches test requests
to another NHS website, these data are less likely to reflect testing
demands. Only the last 7 weeks of data are used.

**Note**: this research has not been peer-reviewed yet. This website is still
experimental. Please contact the <a
href="mailto:thibautjombart@gmail.com">authors</a> before using its content.


Models considered here for trend fitting include:

* linear regression with a linear time effect
* Poisson GLMs
    + constant
    + with log-linear time effect
    + with log-linear time effect and weekend effect (weekend / monday / other
      day)
    + with log-linear time effect and weekday effect
* Negative Binomial GLMs
    + constant
    + with log-linear time effect
    + with log-linear time effect and weekend effect (weekend / monday / other
      day)
    + with log-linear time effect and weekday effect

Analyses are run separately for each NHS region.

```{r packages}
library(incidence2)
library(trendbreaker)
library(ggplot2)
library(cowplot)
library(dplyr)
library(magrittr)
```


```{r prep}
# get the latest data
folder <- file.path("data", "clean")
filenames <- sort(list.files(folder), decreasing = TRUE)
pathways <- readRDS(file.path(folder, filenames[1]))

# tidy up names
pathways$ccg_name <- sub("_ccg$", "", pathways$ccg_name)
first_date <- max(pathways$date, na.rm = TRUE) - 7*7
pathways_recent <- pathways[pathways$date >= first_date, ]

# define candidate models
models <- list(
  regression = lm_model(count ~ date_index),
  poisson_constant = glm_model(count ~ 1, family = "poisson"),
  poisson_time = glm_model(count ~ date_index, family = "poisson"),
  poisson_time_daytype = glm_model(count ~ day_type + date_index, family = "poisson"),
  poisson_time_weekday = glm_model(count ~ weekday + date_index, family = "poisson"),
  negbin_constant = glm_nb_model(count ~ 1),
  negbin_time = glm_nb_model(count ~ date_index),
  negbin_time_daytype = glm_nb_model(count ~ day_type + date_index),
  negbin_time_weekday = glm_nb_model(count ~ weekday + date_index)
)

# function to assign custom day of the week to dates
day_of_week <- function(date) {
  day_of_week <- weekdays(date)
  out <- vapply(
    day_of_week,
    function(x) {
      if (x %in% c("Saturday", "Sunday")) {
        "Weekend"
      } else if (x == "Monday") {
        "Monday"
      } else {
        "rest_of_week"
      }
    },
    character(1)
  )
  factor(out, levels = c("rest_of_week", "Monday", "Weekend"))
} 
```


## ASMODEE results

### England

```{r }

# generate counts for all data
counts_england <- pathways_recent %>%
  filter(site_type == "111_online") %>% 
  incidence(date_index = date,
            count = count
)

# add on day and weekday
counts_england <- counts_england %>%
  mutate(day_type = day_of_week(date_index),
         weekday = weekdays(date_index))

## results with 'k' fixed to 7 days
res_england <- asmodee(
  counts_england,
  models,
  method = evaluate_aic,
  alpha = 0.05,
  fixed_k = 7
)

plot(res_england) +
  theme(text = element_text(size = 16),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_date(date_label = format("%d %b")) +
  labs(x = NULL,
       y = "Daily, potential, COVID-19 reports")

```



### NHS regions

```{r results_nhs_region, fig.height = 7}

# generate for regions
counts_nhs_region <- incidence(
  pathways_recent,
  date_index = date,
  groups = nhs_region,
  count = count,
  na_as_group = FALSE
)


# add on day and weekday
counts_nhs_region <- counts_nhs_region %>%
  mutate(day_type = day_of_week(date_index),
         weekday = weekdays(date_index))

res_nhs_region <- asmodee(
  counts_nhs_region,
  models,
  method = evaluate_aic,
  alpha = 0.05,
  fixed_k = 7
)

plot(res_nhs_region)

```




## Analyses by age groups

This section reproduces the same analyses as above, broken down by age groups.

### All of England

```{r }

# generate for regions
counts_england_age <- pathways_recent %>%
  mutate(age = if_else(age == "70-120", "70+", age)) %>%
  incidence(
      date_index = date,
      groups = age,
      count = count,
      na_as_group = FALSE
  )

# add on day and weekday
counts_england_age <- counts_england_age %>%
  mutate(day_type = day_of_week(date_index),
         weekday = weekdays(date_index))

# run asmodee
res_england_age <- asmodee(
  counts_england_age,
  models,
  method = evaluate_aic,
  alpha = 0.05,
  fixed_k = 7
)

```

```{r echo = FALSE}

n_figs <- length(res_england_age)

```

```{r fig.height = 2 * round(n_figs / 3), out.width = "100%", dpi = 70} 

res_england_age <- res_england_age[sort(names(res_england_age))]
plot(res_england_age, ncol = 3)

```


### NHS regions

```{r }

# generate for regions
counts_nhs_region_age <- pathways_recent %>%
  mutate(age = if_else(age == "70-120", "70+", age)) %>%
  incidence(
      date_index = date,
      groups = c(nhs_region, age),
      count = count,
      na_as_group = FALSE
  )

# add on day and weekday
counts_nhs_region_age <- counts_nhs_region_age %>%
  mutate(day_type = day_of_week(date_index),
         weekday = weekdays(date_index))

# run asmodee
res_nhs_region_age <- asmodee(
  counts_nhs_region_age,
  models,
  method = evaluate_aic,
  alpha = 0.05,
  fixed_k = 7
)
```

```{r echo = FALSE}

n_figs <- length(res_nhs_region_age)

```

```{r fig.height = 2 * round(n_figs / 3), out.width = "100%", dpi = 70} 

res_nhs_region_age <- res_nhs_region_age[sort(names(res_nhs_region_age))]
plot(res_nhs_region_age, ncol = 3)

```


