---
title: "Covid-19 trends in England"
description: |
  Using ASMODEE to monitor NHS pathways in England
author:
  - name: Thibaut Jombart, Dirk Schumacher 
date: "`r Sys.Date() - 1`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = "hide", fig.height = 5, fig.width = 8, dpi = 80, layout="l-body-outset")
```

Using the new algorithm ASMODEE (**A**utomatic **S**election of **M**odels and
**O**utlier **D**etection for **E**pidemics) we monitor changes in potential
COVID-19 cases reported through the [NHS
pathways](https://digital.nhs.uk/dashboards/nhs-pathways), including calls to
111, 999, and 111-online. Only the last 6 weeks of data are used.

**Note**: this research has not been peer-reviewed yet. This website is still
experimental. Please contact the <a
href="mailto:thibautjombart@gmail.com">authors</a> before using its content.


Models considered here for trend fitting include:

1. linear regression of counts over time

2. Poisson GLM with a constant rate

3. Poisson GLM with a log-linear effect of time

4. Negative Binomial GLM with a log-linear effect of time

5. Negative Binomial GLM with a log-linear effect of time and a 'weekday' effect
distinguishing weekends, Mondays, and the rest of the week


Analyses are run separately for each NHS region.

```{r packages}
library(incidence2)
library(trendbreaker)
library(ggplot2)
library(cowplot)
```


```{r prep}
# get the latest data
folder <- file.path("data", "clean")
filenames <- sort(list.files(folder), decreasing = TRUE)
pathways <- readRDS(file.path(folder, filenames[1]))

# tidy up names
pathways$ccg_name <- sub("_ccg$", "", pathways$ccg_name)
first_date <- max(pathways$date, na.rm = TRUE) - 6*7
pathways_recent <- pathways[pathways$date >= first_date, ]

# define candidate models
models <- list(
  regression = lm_model(count ~ day),
  poisson_constant = glm_model(count ~ 1, family = "poisson"),
  poisson_time = glm_model(count ~ day, family = "poisson"),
  poisson_time_weekday = glm_model(count ~ day + weekday, family = "poisson"),
  negbin_time = glm_nb_model(count ~ day),
  negbin_time_weekday = glm_nb_model(count ~ day + weekday)
)

# function to assign custom day of the week to dates
day_of_week <- function(date) {
  day_of_week <- weekdays(date)
  out <- vapply(
    day_of_week,
    function(x) {
      if (x %in% c("Saturday", "Sunday")) {
        "Weekend"
      } else if (x == "Monday") {
        "Monday"
      } else {
        "rest_of_week"
      }
    },
    character(1)
  )
  factor(out, levels = c("rest_of_week", "Monday", "Weekend"))
} 
```

## England

```{r results_england}

# generate counts for all data
counts_england <- incidence(
  pathways_recent,
  date_index = date,
  count = count
)

# add on day and weekday
counts_england$day <- as.integer(
  counts_england$date - min(counts_england$date, na.rm = TRUE)
)
counts_england$weekday <- day_of_week(counts_england$date)
         
## results with 'k' fixed to 7 days
res_england <- asmodee(
  counts_england,
  models,
  method = evaluate_aic,
  alpha = 0.05,
  fixed_k = 7
)

plot(res_england, "date") +
  theme(text = element_text(size = 16),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_date(date_label = format("%d %b")) +
  labs(x = NULL,
       y = "Daily, potential, COVID-19 reports")

```



## NHS regions

```{r results_nhs_region, fig.height = 7}

# generate for regions
counts_nhs_region <- incidence(
  pathways_recent,
  date_index = date,
  groups = nhs_region,
  count = count,
  na_as_group = FALSE
)

# add on day and weekday
counts_nhs_region$day <- as.integer(
  counts_nhs_region$date - min(counts_nhs_region$date, na.rm = TRUE)
)
counts_nhs_region$weekday <- day_of_week(counts_nhs_region$date)

res_nhs_region <- asmodee(
  counts_nhs_region,
  models,
  method = evaluate_aic,
  alpha = 0.05,
  fixed_k = 7
)

plots_nhs_region <- lapply(seq_along(res_nhs_region),
                           function(i)
                             plot(res_nhs_region[[i]], "date", point_size = 1, guide = FALSE) +
                               theme(text = element_text(size = 12),
                                     axis.text.x = element_text(angle = 45, hjust = 1)) +
                               scale_x_date(date_label = format("%d %b")) +
                               labs(x = NULL,
                                    y = NULL,
                                    subtitle = names(res_nhs_region)[i]))
plot_grid(plotlist = plots_nhs_region)
```
