---
title: "Covid-19 trends by CCG level"
description: |
  Using ASMODEE to monitor NHS pathways by CCG
author:
  - name: Thibaut Jombart, David Simons, Dirk Schumacher, Tim Taylor
date: "`r Sys.Date() - 1`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = "hide", fig.height = 5, fig.width = 8, dpi = 80)
```

Using the new algorithm ASMODEE (**A**utomatic **S**election of **M**odels and
**O**utlier **D**etection for **E**pidemics) we monitor changes in potential
COVID-19 cases reported through the [NHS
pathways](https://digital.nhs.uk/dashboards/nhs-pathways), including calls to
111, 999, and 111-online. These analyses are broken down by Clinical
Commissioning Groups [(CCG)](https://www.england.nhs.uk/ccgs/). Only the last 6
weeks of data are used. The last week of data is not used to define the temporal
trend so that recent outliers can be detected.

**Note**: this research has not been peer-reviewed yet. This website is still
experimental. Please contact the <a
href="mailto:thibautjombart@gmail.com">authors</a> before using its content.


* Gaussian GLM and Negative Binomial GLMs
* With the following effects
    + constant, linear, or log-linear time effect
    + optional: weekend effect (weekend / monday / other
      day, or weekday effect
	  )
    + optional: with a change in slope at a given day, from day 9 to day 25
	
Analyses are run separately for each CCG.

```{r packages}
library(incidence2)
library(trendbreaker)
library(ggplot2)
library(cowplot)
library(DT)
library(dplyr)
library(tibble)
library(tidyr)
library(sf)
library(i2extras)
library(leaflet)
library(leafpop)
```

```{r prep}
# get the latest data
folder <- file.path("data", "clean")
filenames <- sort(list.files(folder), decreasing = TRUE)
pathways <- readRDS(file.path(folder, filenames[1]))

# tidy up names
pathways$ccg_name <- sub("_ccg$", "", pathways$ccg_name)
first_date <- max(pathways$date, na.rm = TRUE) - 7 * 7
pathways_recent <- pathways[pathways$date >= first_date, ]

# function to assign custom day of the week to dates
day_of_week <- function(date) {
  day_of_week <- weekdays(date)
  out <- vapply(
    day_of_week,
    function(x) {
      if (x %in% c("Saturday", "Sunday")) {
        "Weekend"
      } else if (x == "Monday") {
        "Monday"
      } else {
        "rest_of_week"
      }
    },
    character(1)
  )
  factor(out, levels = c("rest_of_week", "Monday", "Weekend"))
}

# function to convert character string to snake_case
to_snake_case <- function(x) {
  tmp <- gsub(",", "", x)
  tolower(gsub("\\s+", "_", tmp))
}
``` 




```{r }

# get count by CCG
counts_ccg_all <- incidence(
  pathways_recent,
  date_index = date,
  count = count,
  groups = ccg_name
)

tmp <- attributes(counts_ccg_all)

counts_ccg_all <- tidyr::complete(
  counts_ccg_all,
  tidyr::expand(counts_ccg_all,date_index, ccg_name), 
  fill = list(count = 0)
)

new_row_names <- attr(counts_ccg_all, "row.names")
attributes(counts_ccg_all) <- tmp
attr(counts_ccg_all, "row.names") = new_row_names


# add on day and weekday
counts_ccg_all <- counts_ccg_all %>%
  mutate(day_type = day_of_week(date_index),
         weekday = weekdays(date_index),
         day = as.integer(date_index - min(date_index)))

min_k <- 9
max_k <- 25
k_values <- min_k:max_k
change_k_df <- lapply(k_values,
                       function(k)
                         counts_ccg_all %>%
                         transmute(if_else(day <= k, "before", "after")) %>%
                         pull(1)) %>%
  data.frame() %>%
  tibble() %>%
  setNames(paste0("change_", k_values))

counts_ccg_all <- counts_ccg_all %>%
  bind_cols(change_k_df)


# generate models
model_grid <- expand.grid(
    "date_index", # time effect
    c("", "day_type", "weekday"), # optional offsets for specific days
    c("", paste("date_index*change", k_values, sep = "_")) # optional split
)

## convert to text
predictors_txt <- model_grid %>%
  apply(1, paste, collapse = " + ")

## cleanup
predictors_txt <- gsub("(\\+[ ]*)+[ ]*\\+", " + ", predictors_txt) # +... + -> +
predictors_txt <- sub("^[ ]*\\+", "", predictors_txt) # heading +
predictors_txt <- sub("\\+[ ]*$", "", predictors_txt) # trailing +
predictors_txt <- sub("^[ ]+", "", predictors_txt) # heading spaces
predictors_txt <- sub("[ ]+$", "", predictors_txt) # trailing spaces
predictors_txt <- sub("[ ]+", " ", predictors_txt) # multiple spaces

## add constant models
predictors_txt <- c("1", predictors_txt)



models_txt  <- c(
    sprintf("glm_model(count ~ %s, family = 'gaussian')", predictors_txt), # Gaussian GLMs
    sprintf("glm_nb_model(count ~ %s)", predictors_txt) # NegBin GLMs
)

length(models_txt)
head(models_txt)
tail(models_txt)

models <- lapply(models_txt, function(e) eval(parse(text = e)))

```






## Results: 111/999 calls and 111-online.

These analyses use calls to 111 and 999, as well as NHS 111-online reports of
potential COVID-19 cases.

```{r results_ccg_calls}

## run asmodee on all CCGs
## note: this takes about 1 minute to run with AIC model selection,
## around 16-17 min with cross validation
res_ccg_all <- asmodee(
    counts_ccg_all,
    models,
    method = evaluate_aic,
    fixed_k = 7,
    alpha = 0.05
)

## ## For loop, useful to debug if specific ccgs fail
## for (i in 1:length(counts_ccg_all)) {
##   asmodee(counts_ccg_all[[i]],
##           models,
##           method = evaluate_aic,
##           fixed_k = 7,
##           alpha = 0.05)
## }

## derive summary stats
ccg_all_stats <- 
  summary(res_ccg_all) %>% 
  arrange()
ccg_all_stats <- summary(res_ccg_all)
ccg_all_stats$last_pred <- vapply(
  res_ccg_all, 
  function(x) tail(x$results$estimate, 1),
  double(1)
)
ccg_all_stats <- 
  rename(ccg_all_stats, ccg = group) %>% 
  arrange(ccg_all_stats, desc(n_recent_increases), desc(last_pred)) %>%
  tibble()

```



### Trend changes over the last week: overview

This graph provides an overview of the numbers of increases/decreases detected
at a CCG level by ASMODEE over the last week.

```{r ccg_outliers_summary, out_width = "50%"}

ccg_outliers_summary <- ccg_all_stats %>%
  mutate(n_recent_increases = factor(n_recent_increases, levels = 0:7),
         n_recent_decreases = factor(n_recent_decreases, levels = 0:7)) %>%
  group_by(n_recent_increases, n_recent_decreases) %>%
  count() %>%
  rename(increase = n_recent_increases,
         decrease = n_recent_decreases,
         frequency = n) %>%
  pivot_longer(1:2, names_to = "change", values_to = "n")

ccg_outliers_summary  %>%
  ggplot(aes(x = n, y = frequency, fill = change)) +
  geom_col(position = "dodge") +
  scale_fill_manual("Change over last week",
                    values = c(increase = "#B26363", decrease = "#93bca8")) +
  theme_bw() +
  labs(x = "Number of days showing trend change last week",
       y = "Number of CCGs",
       title = "Trend changes detected by ASMODEE") +
  theme(legend.position = "bottom")

```


```{r echo = FALSE}

n_figs <- length(res_ccg_all)

```

```{r reorder_ccg} 

## reorder ccgs
ccg_order <- ccg_all_stats %>%
  filter(ccg != "null") %>% 
  arrange(desc(n_recent_increases),
          desc(n_recent_outliers)) %>%
  pull(ccg)

oclass <- class(res_ccg_all)
res_ccg_all <- res_ccg_all[ccg_order]
names(res_ccg_all) <- sub("nhs_", "", names(res_ccg_all))
class(res_ccg_all) <- oclass
```


```{r ccg_map, echo = FALSE}
ccg_shape <- st_read(file.path("data", "shapefile", "ccg_2021.shp"))

# create same matching names for ccg
ccg_shape_aligned <- ccg_shape %>%
  mutate(
    ccg = to_snake_case(CCG21NM),
    ccg = gsub("_ccg", "", ccg, fixed = TRUE),
    ccg = recode(ccg, "nhs_central_london_westminster" = "nhs_central_london_(westminster)") 
  ) 

# make plots
individual_plots <- lapply(res_ccg_all, plot, guide = FALSE)

# calculate growth for ccgs
r_ccg <- counts_ccg_all %>%
  keep_last(3*7) %>%
  fit_curve("poisson") %>%
  growth_rate() %>%
  rename("ccg" = "ccg_name")

CCG <- ccg_shape_aligned %>%
  full_join(., ccg_all_stats, by = "ccg") %>%
  full_join(., r_ccg, by = "ccg") %>%
  filter(ccg != "null") %>% # there is a single ccg set as null I've removed it #
  select(CCG21NM, ccg, n_recent_outliers, n_recent_increases, n_recent_decreases, r, r_lower, r_upper, geometry) %>%
  mutate(
    n_recent_increases = factor(n_recent_increases),
    n_recent_decreases = factor(n_recent_decreases),
    confidence = ifelse(r_lower > 0 & r_upper > 0, "confident_growth",
                        ifelse(r_lower < 0 & r_upper < 0, "confident_decay",
                               "equivocal")),
    ccg = gsub("nhs_", "", ccg, fixed = TRUE),
    plots = individual_plots[ccg]
  ) %>% 
  rename( # tidying the column titles for mapview
    `CCG Name` = CCG21NM,
    `Number of outlier days (last week)` = n_recent_outliers,
    `Days increased` = n_recent_increases,
    `Number of days decreased` = n_recent_decreases
  ) %>%
  st_transform(., "WGS84")
```



### CCGs with number of ASMODEE outlier days and growth rate

Selecting the outlier radial will display the number of days each CCG is outlying
the ASMODEE predictions for, the growth rate radial displays the CCG growth rate. 
Selecting the CCG of interest will present the ASMODEE plot for that CCG

The growth rate (*r*) can be roughly interepreted as the percentage of daily
increase / decrease in case incidence. For instance, a *r* of 0.015 means
incidence increases by 1.5% every day, while a value of -0.02 means incidence
decreases by approximately 2% every day. See this excellent
[article](https://plus.maths.org/content/epidemic-growth-rate) for more
information on *r* and its relationship with the reproduction number, *R*.

``` {r combined, results="markup"}
CCG <- select(CCG, -ccg)
plots <- CCG$plots

# Set the palette explicitly as I want 0 to have a high alpha
factpal <- colorFactor(palette = "Blues", CCG$`Days increased`, alpha = T) 
contpal <- colorNumeric(palette = "RdYlBu",  domain = CCG$r, reverse = T)

leaflet(CCG,
        height = 620) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(color = "black",
              fillColor = ~factpal(`Days increased`), 
              fillOpacity = 0.9, 
              weight = 1,
              smoothFactor = 0.2,
              label = CCG$`CCG Name`,
              popup = popupGraph(plots,
                                 width = 260,
                                 height = 180),
              highlightOptions = highlightOptions(color = "black", weight = 2,
                                                  bringToFront = TRUE),
              group = "Outliers") %>%
  addLegend("topright",
            title = "Days increased",
            pal = factpal,
            values = ~`Days increased`,
            group = "Outliers") %>%
  addPolygons(color = "black",
              fillColor = ~contpal(r), 
              fillOpacity = 0.9, 
              weight = 1,
              smoothFactor = 0.2,
              label = CCG$`CCG Name`,
              popup = popupGraph(plots,
                                 width = 260,
                                 height = 180),
              highlightOptions = highlightOptions(color = "black", weight = 2,
                                                  bringToFront = TRUE),
              group = "Growth rate") %>%
  addLegend("topright",
            title = "Growth rate",
            pal = contpal,
            values = ~r,
            group = "Growth rate") %>%
  addLayersControl(
    baseGroups = c("Outliers", "Growth rate"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  hideGroup("Growth rate")
```

### CCG by numbers of outliers

The following graph shows ASMODEE results ordered by decreasing amounts of
outliers in the last 7 days.

```{r plots_all, eval = TRUE, fig.height = 2 * round(n_figs / 3), out.width = "100%", dpi = 50} 
# plot results for top ccg
plot(res_ccg_all, ncol = 3)

```

### Table summary for all CCGs

This table summarises the numbers of changes detected by ASMODEE over the last 7
days, broken down into increases and decreases. *prob* is the probability of
observing that many outliers in the last week under the best fitting model,
given the chosen alpha threshold (5%).

```{r ccg_all_table, results = "markup"}

# display table
ccg_all_stats %>%
  mutate(prob = format.pval(p_recent_outliers, digits = 3),
         ccg = gsub("_", " ", ccg),
         ccg = gsub("nhs", "NHS", ccg)) %>%
  select(ccg,
         changes = n_recent_outliers,
         increase = n_recent_increases,
         decrease = n_recent_decreases,
         prob) %>%
  DT::datatable(ccg_all_stats,
                rownames = FALSE,
                width = "100%",
                class = 'cell-border stripe',
                filter = 'top')
```
