---
title: "COVID-19 trends by CCG level"
author: "Thibaut Jombart, Dirk Schumacher"
date: "`r Sys.Date() - 1`"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      results = "hide",
                      collapse = TRUE,
                      dpi = 80,
                      fig.height = 5,
                      fig.width = 8,
                      out.width = "100%",
                      warning = FALSE,
                      message = FALSE,
                      dev = "png")
```

Using the new algorithm ASMODEE (**A**utomatic **S**election of **M**odels and
**O**utlier **D**etection for **E**pidemics) we monitor changes in potential
COVID-19 cases reported through the [NHS
pathways](https://digital.nhs.uk/dashboards/nhs-pathways), including calls to
111, 999, and 111-online. These analyses are broken down by Clinical
Commissioning Groups [(CCG)](https://www.england.nhs.uk/ccgs/). Only the last 6
weeks of data are used. The last week of data is not used to define the temporal
trend so that recent outliers can be detected.

**Note**: this research has not been peer-reviewed yet. This website is still
experimental. Please contact the <a
href="mailto:thibautjombart@gmail.com">authors</a> before using its content.


Models considered here for trend fitting include:

* linear regression of counts over time

* Poisson GLM with a constant rate

* Negative Binomial GLM with a log-linear effect of time

* Negative Binomial GLM with a log-linear effect of time and a 'weekday' effect
distinguishing weekends, Mondays, and the rest of the week

Analyses are run separately for each CCG.


```{r installs}

Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS = "true", eval = FALSE)
repo <- "https://cloud.r-project.org/"

if(!require(tidyverse)) {
  install.packages("tidyverse", repos = "https://cloud.r-project.org/")
  library(tidyverse)
}

if(!require(remotes)) {
  install.packages("devtools", repos = "https://cloud.r-project.org/")
  library(devtools)
}

if(!require(DT)) {
  install.packages("DT", repos = "https://cloud.r-project.org/")
  library(DT)
}

if(!require(incidence2)) {
  install.packages("incidence2", repos = "https://cloud.r-project.org/")
  library(incidence2)
}

remotes::install_github("reconhub/trendbreaker", upgrade = FALSE)
library(trendbreaker)

```



```{r prepare_data}

# download data
pathways <- tempfile()
download.file("https://github.com/qleclerc/nhs_pathways_report/raw/master/data/rds/pathways_latest.rds", pathways)
pathways <- readRDS(pathways)

## No longer used, but left as backup
## # function to assign custom day of the week to dates
day_of_week <- function(date) {
  day_of_week <- weekdays(date)
  out <- dplyr::case_when(
    day_of_week %in% c("Saturday", "Sunday") ~ "weekend",
    day_of_week %in% c("Monday") ~ "monday",
    TRUE ~ "rest_of_week"
  )
  out <- factor(out, levels = c("rest_of_week", "monday", "weekend"))
  out
}

pathways <- as_tibble(pathways) %>%
  mutate(ccg_name = sub("_ccg$", "", ccg_name))

## last_date <- as.Date("2020-06-15")
last_date <- max(pathways$date, na.rm = TRUE)
first_date <- last_date - 6*7
pathways_recent <- pathways %>%
  filter(date >= first_date,
         date <= last_date)

pathways_recent_calls <- pathways_recent %>%
  filter(site_type %in% c("111", "999"))



# define candidate models
models <- list(
  regression = lm_model(count ~ day),
  poisson_constant = glm_model(count ~ 1, family = "poisson"),
  ## quasipoisson_time = glm_model(count ~ day, family = "quasipoisson"),
  ## quasipoisson_time_weekday = glm_model(count ~ day + weekday, family = "quasipoisson"),
  ## quasipoisson_time_weekday2 = glm_model(count ~ day * weekday, family = "quasipoisson")
  negbin_time = glm_nb_model(count ~ day,
                             control = glm.control(maxit = 1e3)),
  negbin_time_weekday = glm_nb_model(count ~ day + weekday,
                                     control = glm.control(maxit = 1e3))
  )


 

```






## Results: 111/999 calls and 111-online.

These analyses use calls to 111 and 999, as well as NHS 111-online reports of
potential COVID-19 cases. Only CCGs with at least 10 cases in the training set
are used in the analyses, as time series with lower counts currently cause
errors in some of the model candidates. This will be addressed in further
improvements to the [*trending*](https://github.com/reconhub/trending/issues/10)
package.

```{r results_ccg_calls}

# analyses by CCG

## get count by CCG
counts_ccg_all <- pathways_recent %>%
  incidence(date = date, groups = ccg_name, count = count) %>%
  mutate(day = as.integer(date - min(date)),
         weekday = day_of_week(date)) %>%
  split(.$ccg_name) %>%
  lapply(as.data.frame)

## remove all CCGs with no cases in training set; this is due to the current
## impossibility of negbin GLM to handle these cases, but we are working on a
## fix so that some of the models in the model candidates can fail on the data
## and ASMODEE still work
n_cases_in_training <- counts_ccg_all %>%
  lapply(function(e)
    e %>%
      filter(day < 36) %>%
      summarise(n = sum(count))) %>%
  unlist()
to_keep <- n_cases_in_training > 10
counts_ccg_all <- counts_ccg_all[to_keep]

## to keep only e.g. Leicester
## counts_ccg <- counts_ccg[grep("leicester", names(counts_ccg))]

## run asmodee on all CCGs
## note: this takes about 1 minute to run with AIC model selection,
## around 16-17 min with cross validation
res_ccg_all <- lapply(counts_ccg_all,
                        asmodee,
                        models,
                        method = evaluate_aic,
                        fixed_k = 7,
                        alpha = 0.05)

## ## For loop, useful to debug if specific ccgs fail
## for (i in 1:length(counts_ccg_all)) {
##   asmodee(counts_ccg_all[[i]],
##           models,
##           method = evaluate_aic,
##           fixed_k = 7,
##           alpha = 0.05)
## }

## derive summary stats
ccg_all_stats <- res_ccg_all %>%
  lapply(function(e)
    data.frame(
      p_value = e$p_value,
      k = e$k,
      n_outliers_recent = e$n_outliers_recent,
      n_outliers = e$n_outliers,
      n_recent_increase = sum(tail(e$results$classification, 7) == "increase"),
      n_recent_decrease = sum(tail(e$results$classification, 7) == "decrease"),
      last_pred = tail(e$results$pred, 1))
    ) %>%
  bind_rows(.id = "ccg") %>%
  arrange(desc(n_recent_increase),
          desc(last_pred)) %>%
  tibble()

```




### Trend changes over the last week: overview

This graph provides an overview of the numbers of increases/decreases detected
at a CCG level by ASMODEE over the last week.

```{r ccg_outliers_summary}

ccg_outliers_summary <- ccg_all_stats %>%
  mutate(n_recent_increase = factor(n_recent_increase, levels = 0:7),
         n_recent_decrease = factor(n_recent_decrease, levels = 0:7)) %>%
  group_by(n_recent_increase, n_recent_decrease) %>%
  count() %>%
  rename(increase = n_recent_increase,
         decrease = n_recent_decrease,
         frequency = n) %>%
  pivot_longer(1:2, names_to = "change", values_to = "n")

ccg_outliers_summary  %>%
  ggplot(aes(x = n, y = frequency, fill = change)) +
  geom_col(position = "dodge") +
  scale_fill_manual("Change over last week",
                    values = c(increase = "#B26363", decrease = "#93bca8")) +
  theme_bw() +
  labs(x = "Number of days showing trend change last week",
       y = "Number of CCGs",
       title = "Trend changes detected by ASMODEE")

```



### Top 20 CCG

These plots show the 20 CGGs with the most increases in the last 7 days. This
graph is not displayed if there are no with at least one increase
in the last 7 days.

```{r plots_all, eval = TRUE, fig.height = 19, out.width = "100%"} 

## here we can select results to display as we want: based on low p-values, a
## fixed number of outliers, a value of k, ...

## plot results for top ccg
n_display <- 20

top_ccg_all <- ccg_all_stats %>%
  filter(n_recent_increase > 0) %>% 
  pull(ccg) %>%
  head(n_display)

res_ccg_all_top <- res_ccg_all[top_ccg_all]

if (length(res_ccg_all_top)) {
  plots_ccg_all_top <- seq_along(res_ccg_all_top) %>%
    lapply(
      function(i)
        plot(res_ccg_all_top[[i]], "date", point_size = 1, guide = FALSE) +
          theme(text = element_text(size = 12),
                axis.text.x = element_text(angle = 45, hjust = 1)) +
          scale_x_date(date_label = format("%d %b")) +
          labs(x = NULL,
               y = NULL,
               subtitle = names(res_ccg_all_top)[i]))
  cowplot::plot_grid(plotlist = plots_ccg_all_top, ncol = 3)
}

```




### Table summary for all CCGs

This table summarises the numbers of changes detected by ASMODEE over the last 7
days, broken down into increases and decreases. *prob* is the probability of
observing that many outliers in the last week under the best fitting model,
given the chosen alpha threshold (5%).

```{r ccg_all_table, results = "markup"}

## display table
ccg_all_stats %>%
  mutate(prob = format.pval(p_value, digits = 3),
         ccg = gsub("_", " ", ccg),
         ccg = gsub("nhs", "NHS", ccg)) %>%
  select(ccg,
         changes = n_outliers_recent,
         increase = n_recent_increase,
         decrease = n_recent_decrease,
         p_value) %>%
  DT::datatable(ccg_all_stats,
                rownames = FALSE,
                width = "100%",
                class = 'cell-border stripe',
                filter = 'top')

```




