---
title: "COVID-19 trends by CCG level"
author: "Thibaut Jombart, Dirk Schumacher"
date: "`r Sys.Date() - 1`"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      results = "hide",
                      collapse = TRUE,
                      dpi = 80,
                      fig.height = 5,
                      fig.width = 8,
                      out.width = "100%",
                      warning = FALSE,
                      message = FALSE,
                      dev = "png")
```

Using the new algorithm ASMODEE (**A**utomatic **S**election of **M**odels and
**O**utlier **D**etection for **E**pidemics) we monitor changes in potential
COVID-19 cases reported through the [NHS
pathways](https://digital.nhs.uk/dashboards/nhs-pathways), including calls to
111, 999, and 111-online. These analyses are broken down by Clinical
Commissioning Groups [(CCG)](https://www.england.nhs.uk/ccgs/). Only the last 6
weeks of data are used.

**Note**: this research has not been peer-reviewed yet. This website is still
experimental. Please contact the <a
href="mailto:thibautjombart@gmail.com">authors</a> before using its content.


Models considered here for trend fitting include:

* linear regression of counts over time

* Poisson GLM with a constant rate

* Negative Binomial GLM with a log-linear effect of time

* Negative Binomial GLM with a log-linear effect of time and a 'weekday' effect
distinguishing weekends, Mondays, and the rest of the week

Analyses are run separately for each CCG.


```{r installs}

Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS = "true", eval = FALSE)
repo <- "https://cloud.r-project.org/"

if(!require(tidyverse)) {
  install.packages("tidyverse", repos = "https://cloud.r-project.org/")
  library(tidyverse)
}

if(!require(remotes)) {
  install.packages("devtools", repos = "https://cloud.r-project.org/")
  library(devtools)
}

if(!require(DT)) {
  install.packages("DT", repos = "https://cloud.r-project.org/")
  library(DT)
}

remotes::install_github("reconhub/trendbreaker", upgrade = FALSE)
library(trendbreaker)

```



```{r prepare_data}

# download data
pathways <- tempfile()
download.file("https://github.com/qleclerc/nhs_pathways_report/raw/master/data/rds/pathways_latest.rds", pathways)
pathways <- readRDS(pathways)


# add variables and subsets
day_of_week <- function(date) {
  day_of_week <- weekdays(date)
  out <- dplyr::case_when(
    day_of_week %in% c("Saturday", "Sunday") ~ "weekend",
    day_of_week %in% c("Monday") ~ "monday",
    TRUE ~ "rest_of_week"
  )
  out <- factor(out, levels = c("rest_of_week", "monday", "weekend"))
  out
}

pathways <- as_tibble(pathways) %>%
  mutate(nhs_region = str_to_title(gsub("_"," ",nhs_region)),
         nhs_region = gsub(" Of ", " of ", nhs_region),
         nhs_region = gsub(" And ", " and ", nhs_region),
         day = as.integer(date - min(date, na.rm = TRUE)),
         weekday = day_of_week(date),
         ccg_name = sub("_ccg$", "", ccg_name))

first_date <- max(pathways$date, na.rm = TRUE) - 6*7
pathways_recent <- pathways %>%
  filter(date >= first_date)



# define candidate models
models <- list(
  regression = lm_model(count ~ day),
  poisson_constant = glm_model(count ~ 1, family = "poisson"),
  negbin_time = glm_nb_model(count ~ day),
  negbin_time_weekday = glm_nb_model(count ~ day + weekday),
  negbin_time_weekday2 = glm_nb_model(count ~ day * weekday)
  )


```






## CCG level

```{r results_ccg}

## analyses by CCG
## note: this takes about 1 minute to run with AIC model selection,
## areound 16-17 min with cross validation
counts_ccg <- pathways_recent %>%
  group_by(ccg_name, date, day, weekday) %>%
  summarise(count = sum(count)) %>%
  complete(date, fill = list(count = 0)) %>% 
  split(.$ccg_name)

## run asmodee on all CCGs
res_ccg <- lapply(counts_ccg,
                  asmodee,
                  models,
                  method = evaluate_aic,
                  alpha = 0.05)

## derive summary stats
ccg_stats <- lapply(res_ccg, function(e)
  data.frame(
    p_value = e$p_value,
    k = e$k,
    n_outliers_recent = e$n_outliers_recent,
    n_outliers = e$n_outliers)) %>%
  bind_rows(.id = "ccg") %>%
  arrange(desc(n_outliers),
          desc(k))

```




### Top 18 CCG

These plots show the 18 CGGs with the most outliers in the considered time
period. Future versions will provide plots for all CCGs.

```{r plots, eval = TRUE, fig.height = 14, out.width = "100%"} 

## here we can select results to display as we want: based on low p-values, a
## fixed number of outliers, a value of k, ...

## plot results for top ccg
n_display <- 18

top_ccg <- ccg_stats %>%
  pull(ccg) %>%
  head(n_display)

res_ccg_top <- res_ccg[top_ccg]
plots_ccg_top <- lapply(seq_along(res_ccg_top),
                           function(i)
                             plot(res_ccg_top[[i]], "date", point_size = 1, guide = FALSE) +
                               theme(text = element_text(size = 12),
                                     axis.text.x = element_text(angle = 45, hjust = 1)) +
                               scale_x_date(date_label = format("%d %b")) +
                               labs(x = NULL,
                                    y = NULL,
                                    subtitle = names(res_ccg_top)[i]))
cowplot::plot_grid(plotlist = plots_ccg_top, ncol = 3)

```




### Table summary for all CCGs

```{r ccg_table, results = "markup"}

## display table
ccg_stats %>%
  mutate(p_value = format.pval(p_value, digits = 3)) %>% 
  DT::datatable(ccg_stats, rownames = FALSE)

```




```{r, eval = FALSE}
library(ggplot2)

template <- c(
    "### {{nm}}\n",
    "```{r, echo = FALSE, dpi = 40, out.width = '100%'}\n",
    "lapply(res_ccg_split[['{{nm}}']], plot_batch)\n",
    "```\n",
    "\n"
  )

splitting_fac <- rep(1:length(res_ccg), each = 9)[seq_along(res_ccg)]
res_ccg_split <- split(res_ccg, splitting_fac)
res_ccg_split_names <- sapply(res_ccg_split,
                              function(e)
                                paste(head(names(e), 1), tail(names(e), 1), sep = "_"))
names(res_ccg_split) <- res_ccg_split_names

plot_batch <- function(e) {
  all_plots <- lapply(
    seq_along(e),
    function(i)
      plot(e[[i]], 'date', point_size = 1, guide = FALSE) +
        theme(text = element_text(size = 12),
              axis.text.x = element_text(angle = 45, hjust = 1)) +
        scale_x_date(date_label = format('%d %b')) +
        labs(x = NULL,
             y = NULL,
             subtitle = names(e)[i]))
    cowplot::plot_grid(plotlist = all_plots, ncol = 3)
}

plots <- lapply(
  res_ccg_split_names,
  function(nm) knitr::knit_expand(text = template)
)


Put this outside the chunk, right after
# `r knitr::knit(text = unlist(plots))`

```

